---
title: "Panel Data Introduction"
author: "Tobias RÜttenauer"
format: 
  revealjs:
    scrollable: true
    footer: <[Home](/index.html)> 
---

```{r setup, cache=FALSE, include=FALSE}
# options(htmltools.dir.version = FALSE)
# library(knitr)
# opts_chunk$set(
#   fig.align="center",  
#   fig.height=4, #fig.width=6,
#   # out.width="748px", #out.length="520.75px",
#   dpi=300, #fig.path='Figs/',
#   cache=T#, echo=F, warning=F, message=F
#   )
library(tidyverse)
library(hrbrthemes)
library(fontawesome)
library("plm")
```


## This course  {.smaller}

:::: {.columns}

::: {.column width="60%"}
### Outline

1) Introduction Panel Data
2) Variance Components
3) Estimators: FE & RE & Diff-in-Diff
4) Dynamic Diff-in-Diff
5) Fixed Effects Individual Slopes
:::


::: {.column width="40%"}
### Me

`r fa('address-book')` [Tobias Rüttenauer](https://ruettenauer.github.io/)

`r fa('envelope')` [t.ruttenauer@ucl.ac.uk](mailto:t.ruttenauer@ucl.ac.uk)

`r fa('graduation-cap')` Lecturer in Quantitative Social Science at UCL

`r fa('bicycle')` Environmental Sociology
:::
::::



## Further Materials {.smaller}

This material in Handbook format: @Ruttenauer.2024c

[Slides by Josef Brüderl and Volker Ludwig](https://www.ls3.soziologie.uni-muenchen.de/studium-lehre/archiv/teaching-marterials/panel-analysis_april-2019.pdf). See also @Bruderl.2015.


### Books:

* Intuitive: @Allison.2009

* Comprehensive and formal: @Wooldridge.2010

* For R experts: @Croissant.2019

* General introductions to causal estimation techniques: @Angrist.2015, @Cunningham.2021, @Firebaugh.2008, @Huntington-Klein.2021

The books by @Cunningham.2021 ([Link](https://mixtape.scunning.com/)) and @Huntington-Klein.2021 ([Link](https://theeffectbook.net/)) are freely available online!




## Why panel data analysis?

::: {.fragment fragment-index=1}
In empirical social sciences, we are often interested in the **causal research questions**: we want to investigate questions of cause and effect. 

However, randomized controlled trials (RCT) are often infeasible (e.g. effects of education, marriage, pregnancy).
:::

::: {.fragment fragment-index=2}
A potential middle ground between those two approaches: “**compare alike with alike**”. 

With panel data, we observe the same unit (person, region, or country) repeatedly over time. We can then  not only compare two different units to each other. We can also compare a unit in an earlier stage to the same unit in a later stage.
:::


## Panel Data Structure

Usually cross-sectional data is organized as a matrix, where rows represent the observation / individual and the columns hold the variables. In panel data settings, we need to add the dimension of time. There are two ways to do so:

* Long format: $N \times T$ observations (rows), with variables "id" and "time".

* Wide format: $N$ observations, and $T \times K$ variables, which one variable for each time-period.

## Panel Data Structure

Let's have a look at the "Males" data of the `plm` package.
  
```{r}
library("plm")
data("Males")
head(Males[,1:5], n = 16)
```

## Panel Data Structure

Moreover, there two types of panel data:

* Balanced: Contains information for each unit at each time period

* Unbalanced: Some units have missing information at some time periods

```{r}
is.pbalanced(Males, index = c("nr", "year"))
```


## Panel Attrition

![[SOEP Documentation](https://www.diw.de/documents/publikationen/73/diw_01.c.745900.de/diw_ssp0826.pdf)](fig/soep.PNG)



## Panel surveys

### Some examples

:::: {.columns}

::: {.column width="50%"}

* [PSID](https://psidonline.isr.umich.edu/)

* [National Longitudinal Survey of Youth](https://www.nlsinfo.org/content/cohorts/nlsy79)

* [Understanding Society](https://www.understandingsociety.ac.uk/)

* [Millennium Cohort Study](https://cls.ucl.ac.uk/cls-studies/millennium-cohort-study/)

:::

::: {.column width="50%"}

* [SOEP](https://www.diw.de/de/diw_01.c.678568.de/forschungsdatenzentrum_soep.html)

* [Pairfam](https://www.pairfam.de/)

* [SHARE](http://www.share-project.org/home0.html)

* [LISS](https://www.lissdata.nl/)

:::
::::

The [Comparative Panel File](https://www.cpfdata.com/) provides an infrastructure of data harmonisation across various panels. See @Turek.2021.




## Panel transformation

A nice feature of panel data is that we can do some within-person transformation. For instance we can calculate the lags and leads, or the first differences of data. 

::: {.callout-note}
always make sure the data is sorted properly before you do!
:::



## Panel transformation

Person-specific summary values

```{r}
# Order data
Males <- Males[order(Males$nr, Males$year),]

# Person specific means
Males$m_wage <- ave(Males$wage,
                    Males$nr,
                    FUN = function(x) mean(x, na.rm = TRUE))
```



## Panel transformation

Lag and first difference

```{r}
# Order data
Males <- Males[order(Males$nr, Males$year),]

# Lag (last years value)
Males$lag_wage <- ave(Males$wage,
                      Males$nr,
                      FUN = function(x) dplyr::lag(x, n = 1))

# First difference (this years value minus last years value)
Males$fd_wage <- ave(Males$wage,
                      Males$nr,
                      FUN = function(x) x - dplyr::lag(x, n = 1))
```


## Panel transformation

```{r}
head(Males[, c("nr", "year", "wage", "m_wage", "lag_wage", "fd_wage")], n = 16)
```




## References

